---
title: "EDA Course Project 2"
author: "Joel Edwards"
date: "Sunday, February 22, 2015"
output: html_document
---
=============
Note: Some code makes use of examples used in Udacitys Data Analysis with R

Merges the training and the test sets to create one data set.
------------------------------------------------------------
#### 1 - Download file to a directory and unzip contents
```{r}
fileUrl<-"https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
download.file(fileUrl, destfile="C:/Users/Huitzilopoctli/Desktop/EDA-Course-Project-2/EDAProject.zip")
unzip(zipfile="C:/Users/Huitzilopoctli/Desktop/EDA-Course-Project-2/EDAProject.zip",exdir="C:/Users/Huitzilopoctli/Desktop/EDA-Course-Project-2/")
```

#### 2 - Load RDS files into R
```{r}
NEI <- readRDS("C:/Users/Huitzilopoctli/Desktop/EDA-Course-Project-2/EDAProject/summarySCC_PM25.rds")
SCC <- readRDS("C:/Users/Huitzilopoctli/Desktop/EDA-Course-Project-2/EDAProject/Source_Classification_Code.rds")
```

#### 3 - Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? 
* Average emissions have declined in tonnes
* Median emissions declined slightly (not as pronounced as averages) throughout the 9 year period
* Statistical significance is questionable as variances are extremely high
```{r}
## Make a factor with levels
NEI <- transform(NEI, Year = factor(year))

## How many pollutants are there? Found only One
unique(NEI$Pollutant)

## Run a group by summary of statistics to find mean and standard deviation of Emissions by Years. 
## Given the data emissions have average emissions have declined in tonnes
## and median emissions declined slightly throughout the 9 year period
## however variances are extremely high thus not certain if decline in emissions are stat .significant
library(doBy)
## Used Bargraphs to provide visual comparison of (1) average, (2) median and total emissions
xtrial<-summaryBy(Emissions ~ Year, data = NEI, FUN = function(x) { c(mu = mean(x), std = sd(x), md =median(x)) })
xtrial

par(mfrow = c(1, 2))
bp<-barplot(xtrial$Emissions.mu, 
            col = rainbow(20, start = 0, end = 1),
            main = "Average Emissions from 1999 - 2008",
            xlab = "Years",
            ylab = "Average Emissions (Tonnes)")
axis(1,at=bp,labels=xtrial$Year)

boxplot(log(Emissions+1) ~ Year, 
        data=NEI,
        ylim=c(0,1),
        col = rainbow(20, start = 0, end = 1),
        main = "Log(Median) Emissions from 1999 - 2008",
        xlab = "Years",
        ylab = "Log(Median) Emissions (Tonnes)")

```

#### 4 - Have total emissions from PM2.5 decreased in the Baltimore City from 1999 to 2008? 
* Average emissions have declined in tonnes
* Median emissions declined slightly throughout the 9 year period
* Variances have declined. Probably grouping across demographics caused increase in previous variance
```{r}
## Subset required.
NEI.maryland.emissions<-subset(NEI, fips==24510, select=c("Emissions", "type", "Year"))

## Statistical summary of mean, sd, and median.
xtrial<-summaryBy(Emissions ~ Year, data = NEI.maryland.emissions, FUN = function(x) { c(mu = mean(x), std = sd(x), md =median(x)) } )
xtrial

par(mfrow = c(1, 2))
bp<-barplot(xtrial$Emissions.mu, 
            col = rainbow(20, start = 0, end = 1),
            main = "Average Emissions from 1999 - 2008 (Maryland)",
            xlab = "Years",
            ylab = "Average Emissions (Tonnes)")
axis(1,at=bp,labels=xtrial$Year)

boxplot(log(Emissions+1) ~ Year, 
        data=NEI,
        ylim=c(0,1),
        col = rainbow(20, start = 0, end = 1),
        main = "Log(Median) Emissions from 1999 - 2008 (Maryland)",
        xlab = "Years",
        ylab = "Log(Median) Emissions (Tonnes)")
```

#### 5 - Which sources have seen decreases/increases in emissions from 1999-2008 for Baltimore City?
* Statistical summary illustrates declines in average emission rates across all emission types.
* Graphs show outliers influence the averages, thus statistical medians are more representative.

* Using medians and graphs, Road and Non-road emissions have initially declined then held steady.
* Point emissions have declined over time continually.
* Non-point emissions vary, initially increasing in emissions then subsiding down in 2008
```{r}
## Code guidelines adapted from: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
library("ggplot2")
xtrial<-summaryBy(Emissions ~ type + Year, data = NEI.maryland.emissions, FUN = function(x) { c(mu = mean(x), std = sd(x), md =median(x)) } )
xtrial

ggplot(NEI.maryland.emissions, aes(Year, Emissions), main="Emissions by type from 1999 - 2008: Maryland measurements") + geom_point(aes(color = type), size = 5, alpha = 0.5) + facet_grid(. ~type) + stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, geom = "crossbar", width = 0.5) + ylim(0,250)

```
#### 6 - Across the US, how have coal emissions changed between 1999 - 2008?
* Statistical summary illustrates declines in average emission rates across all emission types.
* Graphs show outliers influence the averages, thus statistical medians are more representative.

* Using medians and graphs, Commercial/Insitutional coal emissions have declined (small declines)
* Electric generation shows no change and varies between medians of 1 - 3 tonnes for each examined year.
* Boilers (Industrial) are actually increasing in emissions.
```{r}
## Finds coal sources & respective SCC code
## Derives a temp dataset from NEI where SCC is a coal source 
US.coalsource.emissions <- NEI[NEI$SCC %in% SCC[grep("Coal", SCC$EI.Sector), 1], ]
## For EI Sector associated to SCC code
EISector<-subset(SCC, select=c("SCC","EI.Sector"))
US.coalsource.emissions<-merge(EISector, US.coalsource.emissions, by="SCC")[c(1,2,5,8)]
rm(EISector)

ggplot(US.coalsource.emissions, aes(x = Year, y = Emissions)) + geom_point(aes(color = EI.Sector), size = 10, alpha = 0.3) + facet_grid(. ~EI.Sector) + stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, geom = "crossbar", width = 0.5) + ylim(0,5000)

xtrial<-summaryBy(Emissions ~ EI.Sector + Year, data = US.coalsource.emissions, FUN = function(x) { c(mu = mean(x), std = sd(x), md =median(x)) } )
xtrial
```
#### 7 - How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?
Average and median vehicle emissions have all declined between 1999 - 2008.
```{r}
# Get Baltimore emissions from motor vehicle sources
Maryland.vehilce.emissions <- NEI[NEI$SCC %in% SCC[grep("Vehicle", SCC$EI.Sector), 1], ]
EISector<-subset(SCC, select=c("SCC","EI.Sector"))
Maryland.vehilce.emissions<-merge(EISector, Maryland.vehilce.emissions, by="SCC")
Maryland.vehilce.emissions<-subset(Maryland.vehilce.emissions, fips=="24510", select=c("Emissions", "EI.Sector", "type", "Year", "fips"))

## Used Bargraphs to provide visual comparison of (1) average, (2) median and total emissions
xtrial<-summaryBy(Emissions ~ EI.Sector+Year, data = Maryland.vehilce.emissions, FUN = function(x) { c(mu = mean(x), std = sd(x), md =median(x)) })
xtrial

ggplot(xtrial, aes(x = Year, y = Emissions.mu)) + geom_point(aes(color = EI.Sector), size = 10, alpha = 0.3) + geom_line(aes(group=1)) + facet_grid(. ~EI.Sector)

ggplot(xtrial, aes(x = Year, y = Emissions.md)) + geom_point(aes(color = EI.Sector), size = 10, alpha = 0.3) + geom_line(aes(group=1)) + facet_grid(. ~EI.Sector)
```
#### 8 - Baltimore vs LA Vehicle Emissions?
Greatest change is Baltimore (in terms of change - 300% difference).
In terms of size, LA is the greater.
```{r}
# Prepare LA & Baltimore Data
LA.vehilce.emissions <- NEI[NEI$SCC %in% SCC[grep("Vehicle", SCC$EI.Sector), 1], ]
EISector<-subset(SCC, select=c("SCC","EI.Sector"))
LA.vehilce.emissions<-merge(EISector, LA.vehilce.emissions, by="SCC")
LA.vehilce.emissions<-subset(LA.vehilce.emissions, fips=="06037", select=c("Emissions", "EI.Sector", "type", "Year", "fips"))

LA.Balt.vehicle.emissions<-rbind(Maryland.vehilce.emissions,LA.vehilce.emissions)
LA.Balt.vehicle.emissions$State<-(ifelse(LA.Balt.vehicle.emissions$fips==24510,"Balt","LA"))

## Used Bargraphs to provide visual comparison of (1) average, (2) median and total emissions
xtrial<-summaryBy(Emissions ~ State+Year, data = LA.Balt.vehicle.emissions, FUN = function(x) { c(mu = mean(x), md =median(x), std = sd(x), summation=sum(x))})
xtrial

ggplot(data=xtrial, aes(x=Year, y=Emissions.summation, fill=State)) + 
  geom_bar(stat="identity", position=position_dodge()) + 
  xlab('Year') + 
  ylab('Total Emissions Tonnes') + 
  ggtitle('Comparison of LA vs Baltimore Emissions Over Time' ) 
```



